Data Descriptor for Nature Scientific Data

### 2024-10-30 manuscript/manuscript-nature/manuscript-nature.qmd
res-part-ii

We first focus the analysis on the variability in COVID-19's impact on daily mobility levels across different population groups. Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022. These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset. May 2020 represents the initial phase following the WHO's declaration of COVID-19 as a global pandemic on March 11, 2020, marked by stringent measures. March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis. Figure @Fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index. The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$. Values above this line indicate increased mobility from pre-pandemic levels, while values below show a reduction. These changes show remarkably consistent patterns across countries for both types of classification.

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days. The extent of this decrease tends to be larger for high-density areas. Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density. In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%. However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes. Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions. In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories. Notably, mobility levels in areas characterised by higher population density still remain below the baseline. Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline. These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May). Overall, the observed patterns by population density category are consistent with those reported in a previous work CITE, where the same data was analysed without applying the data processing methodology proposed in this study.

@@

## Initial drop in mobility drives current levels, as recovery pace remains consistent across population groups

Next we focus on

In this section we explore trends in the intensity of movements over time across population groups.
Generally, all categories of analysis experienced an initial drop in the number of movements followed by an upwards general trend of recovery towards baseline levels of mobility.
The extent of the initial drop and the speed of recovery varied by country and by analysis category.
These general patterns can be visually explored in Figure @Fig-recovery-density and @Fig-recovery-rdi for a classification by population density category and by RDI respectively.
Below we further explore these patterns by following a three-step method consisting in i) time series decomposition into general trend, seasonality and noise, ii) modelling the general trend , iii) analysing the factors that influence the values of the parameters estimated in ii) to characterise the general trend.

@@

![](figures/Fig-recovery-rdi.png)
![](figures/Fig-re-density.pdf)

>>>>

We first focus on analysing the variability in COVID-19's impact on daily mobility levels across different population groups. Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022. These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset. May 2020 represents the initial phase following the WHO's declaration of COVID-19 as a global pandemic on March 11, 2020, marked by stringent measures. March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis. Figure @Fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index. The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$. Values above this line indicate increased mobility from pre-pandemic levels, while values below show a reduction. These changes show remarkably consistent patterns across countries for both types of classification.

Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days. The extent of this decrease tends to be larger for high-density areas. Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density. In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%. However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes. Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions. In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories. Notably, mobility levels in areas characterised by higher population density still remain below the baseline. Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline. These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May). Overall, the observed patterns by population density category are consistent with those reported in CITE for Argentina and Chile, where the data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.

@@

## Initial drop in mobility drives intergroup variation in mobility levels, as recovery pace is uniform across population groups

Next we focus on modelling intergroup differences in mobility trends over time towards baseline levels of mobility, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups. Figure @Fig-recovery-density and @Fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively. Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels. The background of each plot in Figures @Fig-recovery-density and @Fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns. This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.

A visual inspection of the evolution patterns by population density category, as shown in @Fig-recovery-density, reveals two main commonalities across the three countries. First, more densely populated areas consistently record lower levels of mobility throughout the study period. Second, there is a general upward trend in mobility across all population density categories for all countries. However, variations exist in the patterns for each country. While Chile experiences greater overall losses in mobility compared to Argentina, certain areas in Argentina reach levels comparable to the pre-pandemic baseline. In contrast, Colombia displays more varied trends across population density categories; low-density areas exhibit above-baseline levels of mobility, whereas high-density areas experience significantly larger losses.

@@

Based on a visual inspection of @Fig-recovery-rdi, the evolution patterns by relative deprivation index category mirror those observed by population density category. First, less deprived areas consistently record lower levels of mobility throughout the study period. In contrast, more deprived areas exhibit higher levels of mobility, particularly in Colombia, where these levels are well above the baseline. However, these significant relative changes do not necessarily correspond to large population numbers. In Colombia, for example, more deprived regions are often low-density areas that typically contain small populations. As a result, minor fluctuations in mobility can lead to substantial percentage changes. Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.

![](figures/Fig-recovery-rdi.png)
Next, we quantitatively model the patterns observed in @Fig-recovery-density and @Fig-recovery-rdi

![](figures/Fig-re-density.pdf)
